### 2012-09-21 doc/prelim/document/chapters/linear_problem/linear_problem.tex
adding msod figure

\chapter{The Linear Problem}
\label{ch:linear_problem}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_finite_2007}, finite volumes
\citep{leveque_finite_2002}, and finite elements
\citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
equations in the form of matrix problems.  In addition, these matrices
are typically sparse, meaning that the vast majority of their
constituent elements are zero. This sparsity is due to the fact that
the influence of a particular grid element only expands as far as a
few of its nearest neighbors depending on the order of discretization
used and therefore coupling among variables in a particular discrete
equation in the system leads to a few non-zero entries. Because of the
natural occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods for
solving sparse systems to provide background and then give detail on a
reemerging stochastic class of methods. Details on the parallelization
of conventional methods are discussed and a novel proposal for the
parallelization of stochastic methods presented.\footnote{The contents
  of this chapter, particularly those sections relating to projection
  methods and matrix analysis, are heavily based on Saad's text
  \citep{saad_iterative_2003}.}

@@

\subsection{Parallel Projection Methods}
\label{subsec:parallel_krylov_methods}
Modern parallel implementations of projection methods on distributed
memory architectures rely heavily on capabilities provided by general
linear algebra frameworks. For methods like GMRES, this arises from
the fact that Krylov methods require only a handful of operation types
in their implementation that can be efficiently programmed on these
architectures. Per Saad's text \citep{saad_iterative_2003} and as
noted in Algorithm~\ref{alg:gmres}, these operations are
preconditioning, matrix-vector multiplications, vector updates, and
inner products. For the last three items, linear algebra libraries
such as PETSc \citep{gropp_scalable_1993} and Trilinos
\citep{heroux_overview_2005} provide efficient parallel
implementations for these operations. Depending on the type of
preconditioning used, efficient parallel implementations may also be
available for those operations. Due to their prevalence in modern
numerical methods, parallel formulations these operations have
warranted intense study \citep{tuminaro_parallel_1998}. In all cases, a series
of scatter/gather operations are required such that global
communication operations must occur. Although the relative performance
of such operations is bound to the implementation, asymptotically
perfomance should be the same across all implementations.

@@

\subsubsection{Parallel Vector Update}
\label{subsubsec:parallel_vec_update}
Parallel vector update operations arise from the construction of the
orthonormal basis and the application of the correction generated by
the constraints to the solution vector. Vector update operations are
embarassingly parallel in that they require no communication
operations to be successfully completed; all data operated on is
local. These operations are globally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_g]
  \:,
  \label{eq:global_vector_update}
\end{equation}
and locally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_l]
  \:,
  \label{eq:local_vector_update}
\end{equation}
where $\ve{y}$ and $\ve{x}$ are vectors of global size $N_g$, local
size $N_l$, and $a \in \mathbb{C}^N$. In order to avoid communcation,
the vectors $\ve{y}$ and $\ve{x}$ must have the same parallel
decomposition where each parallel process owns the same pieces of each
vector.

\subsubsection{Parallel Vector Product}
\label{subsubsec:parallel_vector_product}
Vector product operations are used in several instances during a
Krylov iteration including vector norm computations and the
orthogonalization procedure. By definition, the vector product is a
global operation that effectively collapses two vectors to a single
value. Therefore, we cannot eliminate all global
communications. Instead, vector product operations are formulated as
\textit{global reduction operations} that are efficiently supported by
modern message passing libraries. For the dot product of two vectors
$\ve{y}$ and $\ve{x}$, a single reduction is required such that:
\begin{equation}
  d_l = \ve{y}_l \cdot \ve{x}_l,\ d_g = \sum_p d_l \:,  
  \label{eq:parallel_dot_product}
\end{equation}
where the $l$ subscript denotes a local quantity, $d_l$ is the local
vector dot product, and $d_g$ is the global dot product generated by
summing the local dot products over all $p$ processes. Parallel norm
operations can be conducted with the same single reduction. Consider
the infinity norm operation:
\begin{subequations}
  \begin{gather}
    ||x||_{\infty,l} = \max_n \ve{y}[n],\ \forall n \in [1,N_l]\:\\
    ||x||_{\infty,g} = \max_p ||x||_{\infty,l}\:.
  \end{gather}
  \label{eq:parallel_infinity_norm}
\end{subequations}
In this form, the local infinity norm is computed over the local piece
of the vector. The reduction operation is then formed over all $p$
processes such that the global max of the vector is computed and
distributed to all processes.

\subsubsection{Parallel Matrix-Vector Multiplications}
\label{subsubsec:parallel_mat_vec_mutliply}

@@

\subsubsection{Parallel Perfomance Implications for Krylov Methods}
\label{subsubsec:projection_method_performance}
Knowing the parallel characteristics of the key operations we must
perform in order to implement Krylov methods, we can make a few
statements about parallel perfomance and implications for operation on
machines of increasing size.  For very large distributed machines,
global reduction operations required at several levels of Krylov
algorithms stand to reduce scalablitity and perfomance. Furthermore,
communication between adjacent domains in matrix-vector multiply
operations may also cause a bottleneck as the number of domains used
in simulation grows. The end result is that global data must be
collected and communicated. For scaling improvement, we seek a
reduction in these types of operations. In addition, these issues
become more prominent as the Krylov iterations progress, causing the
Krylov subspace to grow and the total number of operations needed to
orthogonalize that subspace to increase.

\section{Stochastic Methods}
\label{sec:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here
an enjoyable manuscript published in 1950 by Forsythe and Leibler
\citep{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\citep{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\citep{hammersley_monte_1964} presents additional detail on this topic
using a collection of references from the 1950's and early 1960's.

We begin our discussion of stochastic methods using these texts by
seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
operator $\ve{A}$, we can using diagonal splitting in the same manner
as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
define the following operator\footnote{It should be noted that
  non-diagonal splittings have been recently explored in
  \citep{srinivasan_monte_2010} and have the potential to improve
  efficiency.}:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:.
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this inverse sum to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation will therefore lead to an
approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering its $i^{th}$
component:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
in $\ve{H}$ where $\nu$ is a particular random walk sequence
permutation. We can generate these sequences of transitions through
Monte Carlo random walks by assigning them both a probability and
weight. As a reinterpretation of the iteration matrix, we then form
the \textit{Neumann-Ulam decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Hadamard product operation\footnote{The
  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
transition probability matrix, and $\ve{W}$ denotes the transition
weight matrix. This decomposition, a generalization of Dimov's work
\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
scheme in that now a weight cutoff can be used to terminate a random
walk sequence. The formulation of $\ve{P}$ and $\ve{W}$ will be
dependent on whether we choose a direct or adjoint Monte Carlo
sequence to estimate the state transitions in
Eq~(\ref{eq:expanded_neumann_solution}).

\subsection{Direct Method}
\label{subsec:direct_mc}
In the context of matrix inversion, a direct method resembles an
adjoint Monte Carlo method in the reactor physics community where the
solution state is sampled and the source terms that contribute to it
are assembled. To achieve this, we build the direct method
Neumann-Ulam decomposition per Dimov's approach by first choosing a
probability matrix that is a row scaling of $\ve{H}$ such that its
components are:
\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
  \label{eq:direct_probability}
\end{equation}
From this, we then see that the probability of transitioning from a
state $i$ to a state $j$ is implicitly linked to the original operator
$\ve{A}$ in that those terms with large values, and therefore those
that make the greatest contribution to the numerical solution, will be
sampled with a higher probability than smaller terms. In addition, the
row scaling provides a normalization over the state to which we are
transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
the probabilities over the rows of the matrix. The components of
the weight matrix are then defined by
Eq~(\ref{eq:neumann_ulam_decomposition}) as:
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
  \label{eq:direct_weight}
\end{equation}
It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
$\ve{P}$, and $\ve{W}$ must be sparse as well by
definition. Additionally, we only compute $\ve{P}$ and $\ve{W}$ from
the non-zero elements of $\ve{H}$ as those components that are zero
will not participate in the random walk. Furthermore, it prevents an
infinite weight from being generated in Eq~(\ref{eq:direct_weight}).

Using these matrices, we can then form the expectation value of the
direct solution. For a given random walk permutation $\nu$ with $k$
events, we define the weight of that permutation to be:
\begin{equation}
  W_{m} = \sum_{m=0}^k w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
  \label{eq:direct_permutation_weight}
\end{equation}
such that the weight of each transition event contributes to the
total. The contribution to the solution from a particular random walk
permutation is then:
\begin{equation}
  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}\:,
  \label{eq:direct_permutation_contribution}
\end{equation}
where $X_{\nu}(i_0 = i)$ signifies that the solution state in which we
are tallying defines state $i$.  We can interpret this precisely as
before in that during the random walk we collect the source in the
states that are visited and apply them to the solution tally. We then
define the probability that a particular random walk permuation of $k$
events will occur:
\begin{equation}
  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
  \label{eq:direct_permutation_probability}
\end{equation}
Finally, we define the expectation value of $X$ to be the collection
of all random walk permutations and their probabilities:
\begin{equation}
  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{\nu}\:,
  \label{eq:direct_expectation_value}
\end{equation}
which, if expanded, directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X(i_0 = i)\}
    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:direct_expectation_expansion}
\end{equation}
therefore providing an unbiased Monte Carlo estimator.

\subsubsection{Estimator Variance}
\label{subsubsec:estimator_variance}
We can compute the variance of the estimator through traditional
methods by defining the variance, $\sigma_i$, for each component in
the solution:
\begin{equation}
  {\sigma_i}^2 = E\{X(i_0 = i) - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X(i_0
  = i)^2\} - x_i^2\:,
  \label{eq:direct_variance_1}
\end{equation}
where the vector exponentials are computed element-wise. Inserting
Eq~(\ref{eq:direct_expectation_value}) gives:
\begin{equation}
  \sigma_i^2 = \sum_{\nu} P_{\nu} X_{\nu}^2 - x_i^2\:,
  \label{eq:direct_variance_2}
\end{equation}
and applying Eq~(\ref{eq:direct_permutation_contribution}):
\begin{equation}
  \sigma_i^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 -
  x_i^2\:.
  \label{eq:direct_variance_3}
\end{equation}
Finally, expanding the transition probabilities yields:
\begin{equation}
  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_m} - x_i^2\:.
  \label{eq:direct_variance_4}
\end{equation}
Using this defintion for the variance, we can arrive at a more natural
reason for enforcing $\rho(\ve{H}) < 1$. Per the Hadamard product, we can
concatenate the summation in Eq~(\ref{eq:direct_variance_4}):
\begin{equation}
  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
  \label{eq:double_weighted_decomposition}
\end{equation}
If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
formulation for the variance:
\begin{equation}
  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
  x_i^2\:,
\end{equation}
which contains the general Neumann series for $\ve{G}$,
\begin{equation}
  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
  \label{eq:variance_neumann_series}
\end{equation}
where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
into the variance formulation:
\begin{equation}
  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2
  \label{eq:direct_variance_5}
\end{equation}
We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
contains an additional Hadamard product with the weight matrix. The
Hadamard product has the property that:
\begin{equation}
  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:,
  \label{eq:hadamard_inequality}
\end{equation}
meaning that $\rho(\ve{G}) \geq \rho(\ve{H})$. Using these relations
to analyze Eq~(\ref{eq:direct_variance_5}), we see that if
$\rho(\ve{H}) > 1$, then the sum in
Eq~(\ref{eq:variance_neumann_series}) will not converge and an
infinite variance will arise as the elements of $\ve{T}$ become
infinite. Therefore, we must restrict $\ve{H}$ with $\rho(\ve{H}) < 1$
so that our expectation values for the solution may have a finite
variance.

As we seek only approximate solutions, we need only to perform a
minimal number of random walks in order to generate an approximation
for $\ve{x}$. Furthermore, we also need conditions by which we may
terminate a random walk. We do this by noticing that the factors added
to Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
small due to their definition in Eq~(\ref{eq:direct_weight}) such that
their contributions to the solution estimate will become
negligble. Therefore, we choose terminate a random walk sequence with
a \textit{weight cutoff}, $W_c$, that is enforced when $W_m < W_c$ for
a particular random walk permutation.

\subsubsection{Reusing Random Walks}
\label{subsubsec:reusing_random_walks}
Ji recently proposed a variation of the direct method in which the
random walk permutations in Eq~(\ref{eq:mc_walk_permutation}) can be
used to contribute to the solution tally in all visited states instead
of the only starting state \citep{ji_reusing_2012}.

\subsection{Adjoint Method}
\label{subsec:adjoint_mc}
An alternative formulation for Monte Carlo matrix inversion is the
adjoint method. In this scheme, random walk histories are born in a
state generated by source sampling and tallies are generated in the
states in which the random walk currently resides. In terms of
particle transport, this adjoint method is equivalent to a traditional
forward method. To begin, we first form the \textit{adjoint
  Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^T = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where now we are forming the decomposition with respect to the
transpose of $\ve{H}$. We then following the same procedure as the
direct method for forming the probability and weight matrices in the
decomposition. Using the transpose form, probabilities should instead
be column-scaled:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state $j$ from the current state
in the random walk $j$ by sampling column-wise. Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
the weight generated for a particular random walk permutation as in
Eq~(\ref{eq:mc_walk_permutation}) and the fact that in the adjoint
method we are only tallying the solution in the state in which the
walk currently resides, the contribution to the solution from a
particular random walk permutation is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} b_{i_0} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state and $b_{i_0}$ is that sampled source starting
state. Note here that the estimator in
Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
dependency on the source state as in
Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
the situation in the direct method where we must start a random walk
in each source state for every permutation such that we may compute a
contribution for that state. In the adjoint method, we instead tally
in all states and those of lesser importance will not be visited as
frequently by the random walk. Finally, the estimator using all
permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded in the same way as the direct method, directly
recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k} \\ &= x_i\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore also providing an unbiased Monte Carlo estimator.

We also desire a criteria for random walk termination. For the adjoint
method, we utilize a \textit{relative weight cutoff}:
\begin{equation}
  W_f = W_c b_{i_0}\:,
  \label{eq:relative_weight_cutoff}
\end{equation}
where $W_c$ is defined as in the direct method. The adjoint random
walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
contributions become increasingly small.

Given these ideas, we can then generate an algorithm for implementing
the adjoint method:
\begin{algorithm}[htpb!]
  \caption{Adjoint Monte Carlo}
  \label{alg:adjoint_mc}
  \begin{algorithmic}
    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
  \end{algorithmic}
\end{algorithm}

\subsection{Sequential Monte Carlo}
\label{subsec:sequential_mc}
The direct and adjoint methods described are limited by a convergence
rate of $1/\sqrt{N}$ by the Central Limit Theorem where $N$ is the
number of random walk permutations. In 1962, Halton presented a
residual Monte Carlo method that moves towards exponential convergence
rates \citep{halton_sequential_1962} and further refined his work some
years later \citep{halton_sequential_1994}. Applications of his work
by the transport community have confirmed convergence rates on the
order of of $exp(-N)$
\citep{evans_residual_2003}. In
much the same way as projection methods, Halton's method, sequential
Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
driving down the residual vector. He proposed the following iterative
scheme as a solution to Eq~(\ref{eq:linear_problem})\:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{A}\boldsymbol{\delta}^{k+1} = \ve{r}^{k}\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k+1}\:,
  \end{gather}
  \label{eq:sequential_monte_carlo}
\end{subequations}
where the correction $\boldsymbol{\delta}$ is computed by the adjoint
Monte Carlo method. The merits of Halton's approach are immediately
visible in that we have now broken the binding of the convergence rate
to the Central Limit Thereom. Here, the Monte Carlo solve is used to
produce a correction from the residual, analagous to using the
residual to extract a correction from the search subspace in a
projection method. By doing this, the Monte Carlo error is bound in
the correction used to update the solution and therefore does not
explicitly manifest itself in the solution. The downside of such a
method is that if the solution guess is poor, then many iterations are
required in order to reach exponential converge as the Monte Carlo
error (and therefore the Central Limit Theorem) does dominate in this
situation. Therefore, if a good initial guess is not available,
Halton's method is still characterized by poor performance.

\subsection{Monte Carlo Synthetic-Acceleration}
\label{subsec:mcsa}
Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution that was not prohibited severely by the quality of the
initial guess for the system \citep{evans_monte_2009}. Their approach
was instead to use residual Monte Carlo as a synthetic acceleration
for a stationary method. To derive this method, we begin by splitting
the operator in Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
  \label{eq:linear_split}
\end{equation}
With this we can then define the stationary method
\textit{Richardson's iteration} as:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
  \label{eq:richardsons_iteration}
\end{equation}
which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
solution error at the $k^{th}$ iterate relative to the true solution:
\begin{equation}
  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
  \label{eq:mcsa_error}
\end{equation}
Subtracting Eq~(\ref{eq:richardsons_iteration}) from
Eq~(\ref{eq:linear_split}) we get:
\begin{equation}
  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
  \label{eq:mcsa_setup_1}
\end{equation}
Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
\begin{equation}
  \begin{split}
    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
    \label{eq:mcsa_setup_2}
  \end{split}
\end{equation}
Using this, we define the following scheme that will converge in one
iteration if $\ve{A}$ is inverted exactly.
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_setup_3}
\end{subequations}
However, $\ve{A}$ is only approximately inverted by our numerical
methods and therfore we instead pose an iterative scheme in which the
Monte Carlo solvers are used to invert the operator. The
\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
    \label{eq:mcsa}
  \end{gather}
\end{subequations}
where the adjoint Monte Carlo method is used to generate the solution
correction from the residual. Using Monte Carlo in this way achieves
the same effect as Halton's method, decoupling its convergence rate
from the overall convergence rate of the method. Here, the approximate
Monte Carlo solution is not driven to a particular convergence as it
merely supplies a correction for the initial guess generated by
Richardson's iteration. Rather, only a set number of histories are
required using the adjoint method to generate the correction.

In addition to the Monte Carlo solver parameters dictating the number
of histories and weight cutoff, the outer MCSA iterations also have
the following stopping criteria:
\begin{equation}
  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
  \label{eq:mcsa_stopping_criteria}
\end{equation}
where $\epsilon$ is a user-defined parameter. We therefore have 3
parameters to tune in an MCSA implementation: the number of Monte
Carlo histories computed in the adjoint solve during each MCSA
iteration, the weight cutoff for those histories, and the total MCSA
convergence tolerance as specified by $\epsilon$.

>>>>

\chapter{Conventional Solution Methods for Linear Problems}
\label{ch:linear_problem}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_finite_2007}, finite volumes
\citep{leveque_finite_2002}, and finite elements
\citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
equations in the form of matrix problems.  In addition, these matrices
are typically sparse, meaning that the vast majority of their
constituent elements are zero. This sparsity is due to the fact that
the influence of a particular grid element only expands as far as a
few of its nearest neighbors depending on the order of discretization
used and therefore coupling among variables in a particular discrete
equation in the system leads to a few non-zero entries. Because of the
natural occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods for
solving sparse systems to provide the necessary background for the
remainder of this work. Details on the parallelization of conventional
methods are discussed.\footnote{The contents of this chapter,
  particularly those sections relating to projection methods and
  matrix analysis, are heavily based on Saad's text
  \citep{saad_iterative_2003}.}

@@

\section{Parallel Projection Methods}
\label{sec:parallel_krylov_methods}
Modern parallel implementations of projection methods on distributed
memory architectures rely heavily on capabilities provided by general
linear algebra frameworks. For methods like GMRES, this arises from
the fact that Krylov methods require only a handful of operation types
in their implementation that can be efficiently programmed on these
architectures. Per Saad's text \citep{saad_iterative_2003} and as
noted in Algorithm~\ref{alg:gmres}, these operations are
preconditioning, matrix-vector multiplications, vector updates, and
inner products. For the last three items, linear algebra libraries
such as PETSc \citep{gropp_scalable_1993} and Trilinos
\citep{heroux_overview_2005} provide efficient parallel
implementations for these operations. Depending on the type of
preconditioning used, efficient parallel implementations may also be
available for those operations. Due to their prevalence in modern
numerical methods, parallel formulations these operations have
warranted intense study \citep{tuminaro_parallel_1998}. In all cases, a series
of scatter/gather operations are required such that global
communication operations must occur. Although the relative performance
of such operations is bound to the implementation, asymptotically
perfomance should be the same across all implementations.

@@

\subsection{Parallel Vector Update}
\label{subsec:parallel_vec_update}
Parallel vector update operations arise from the construction of the
orthonormal basis and the application of the correction generated by
the constraints to the solution vector. Vector update operations are
embarassingly parallel in that they require no communication
operations to be successfully completed; all data operated on is
local. These operations are globally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_g]
  \:,
  \label{eq:global_vector_update}
\end{equation}
and locally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_l]
  \:,
  \label{eq:local_vector_update}
\end{equation}
where $\ve{y}$ and $\ve{x}$ are vectors of global size $N_g$, local
size $N_l$, and $a \in \mathbb{C}^N$. In order to avoid communcation,
the vectors $\ve{y}$ and $\ve{x}$ must have the same parallel
decomposition where each parallel process owns the same pieces of each
vector.

\subsection{Parallel Vector Product}
\label{subsec:parallel_vector_product}
Vector product operations are used in several instances during a
Krylov iteration including vector norm computations and the
orthogonalization procedure. By definition, the vector product is a
global operation that effectively collapses two vectors to a single
value. Therefore, we cannot eliminate all global
communications. Instead, vector product operations are formulated as
\textit{global reduction operations} that are efficiently supported by
modern message passing libraries. For the dot product of two vectors
$\ve{y}$ and $\ve{x}$, a single reduction is required such that:
\begin{equation}
  d_l = \ve{y}_l \cdot \ve{x}_l,\ d_g = \sum_p d_l \:,  
  \label{eq:parallel_dot_product}
\end{equation}
where the $l$ subscript denotes a local quantity, $d_l$ is the local
vector dot product, and $d_g$ is the global dot product generated by
summing the local dot products over all $p$ processes. Parallel norm
operations can be conducted with the same single reduction. Consider
the infinity norm operation:
\begin{subequations}
  \begin{gather}
    ||x||_{\infty,l} = \max_n \ve{y}[n],\ \forall n \in [1,N_l]\:\\
    ||x||_{\infty,g} = \max_p ||x||_{\infty,l}\:.
  \end{gather}
  \label{eq:parallel_infinity_norm}
\end{subequations}
In this form, the local infinity norm is computed over the local piece
of the vector. The reduction operation is then formed over all $p$
processes such that the global max of the vector is computed and
distributed to all processes.

\subsection{Parallel Matrix-Vector Multiplications}
\label{subsec:parallel_mat_vec_mutliply}

@@

\subsection{Parallel Perfomance Implications for Krylov Methods}
\label{subsec:projection_method_performance}
Knowing the parallel characteristics of the key operations we must
perform in order to implement Krylov methods, we can make a few
statements about parallel perfomance and implications for operation on
machines of increasing size.  For very large distributed machines,
global reduction operations required at several levels of Krylov
algorithms stand to reduce scalablitity and perfomance. Furthermore,
communication between adjacent domains in matrix-vector multiply
operations may also cause a bottleneck as the number of domains used
in simulation grows. The end result is that global data must be
collected and communicated. For scaling improvement, we seek a
reduction in these types of operations. In addition, these issues
become more prominent as the Krylov iterations progress, causing the
Krylov subspace to grow and the total number of operations needed to
orthogonalize that subspace to increase.

### 2012-09-21 doc/prelim/document/chapters/research_proposal/research_proposal.tex
adding msod figure

@@

\section{Key Questions}
\label{sec:key_questions}

>>>>

\section{Experimental Framework}
\label{sec:experimental_framework}

The Chimera code base will be the primary research tool for answering
the questions posed in the previous chapters of the document. We may
not need a chapter for this, but it will likely be a good idea
somewhere in this document to outline the design of the software and
the instrumentation that it will provide.

\subsection{Algorithm-Based Design}
\label{subsec:chimera_design}

\subsection{Graph-Based Multiphysics Formulation}
\label{subsec:multiphysics_graph}

\subsection{Finite Element Assembly}
\label{subsec:fem_assembly}

\subsection{Software Engineering}
\label{subsec:software_engineering}

@@

\section{Numerical Experiments}
\label{sec:key_questions}
